---
title: "Data Sources Overview"
---

# Data Sources Overview

## Accidental spills

### Integrated Satellite Tracking of Pollution (ISTOP) {#sec-istop}

```{r}
#| message: false
istop <- sf::st_read("../../workspace/data/harvested/istop-1.0.0/processed/istop.gpkg", quiet = TRUE)
```

The [Integrated Satellite Tracking of Pollution (ISTOP)](https://www.canada.ca/en/environment-climate-change/services/ice-forecasts-observations/latest-conditions/oil-pollution-monitoring-overview/satellite-spills-overview.html) program [@eccc2024] utilizes satellite imagery to monitor and detect oil spills in coastal waters. Since 2006, daily satellite analyses have been conducted to identify and report both illegal and accidental marine oil pollution. These analyses detect sea-surface anomalies, which are vectorized into polygons representing affected areas. The data obtained span 2006 to 2023 and include `r nrow(istop)` recorded incidents. The data was transferred manually to the Canadian Wildlife Service and are currently stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco.

- **Source**: Environment Canada Integrated Satellite Tracking of Pollution (ISTOP) Program
- **Accessibility**: Restricted
- **Data Type**: Vectorized Polygons
- **Coverage**: North American coastal waters, 2006–2023
- **Processing Script**: `prc_istop.R`
- **Output File**: `istop.gpkg`

### National Aerial Surveillance Program (NASP) {#sec-nasp}

```{r}
#| message: false
nasp <- sf::st_read("../../workspace/data/harvested/nasp-1.0.0/processed/nasp.gpkg", quiet = TRUE)
```

The [National Aerial Surveillance Program (NASP)](https://tc.canada.ca/en/programs/national-aerial-surveillance-program) [@tc2024] is a country-wide aerial surveillance initiative designed to detect pollution and protect Canadian waters. The program, covering pollution detections in the Atlantic region from April 2012 to March 2023, aims to safeguard marine environments, endangered marine life, and promote safe maritime transport along Canada’s coastlines. The dataset includes `r nrow(nasp)` incidents. The data was transferred manually to the Canadian Wildlife Service and are currently stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco.

- **Source**: Transport Canada National Aerial Surveillance Program (NASP)
- **Accessibility**: Restricted
- **Data Type**: CSV
- **Coverage**: Atlantic Region, Canada, 2012–2023
- **Processing Script**: `prc_nasp.R`
- **Output File**: `nasp.gpkg`

### National Environmental Emergency Centre (NEEC) {#sec-neec}

```{r}
#| message: false
neec <- vroom::vroom("../../workspace/data/harvested/neec-1.0.0/processed/neec.csv")
```

The [National Environmental Emergency Centre (NEEC)](https://www.canada.ca/en/environment-climate-change/services/environmental-emergencies-program/national-centre.html) [@neec2024], a program under Environment and Climate Change Canada, provides an inventory of environmental emergency incidents across Canada. The NEEC dataset provided to the Canadian Wildlife Service covers incidents impacting waterbodies in the Atlantic region, including Québec, from 2016 to 2023, comprising `r nrow(neec)` recorded incidents. The data was transferred manually to the Canadian Wildlife Service and are currently stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco.

- **Source**: National Environmental Emergency Centre (NEEC)
- **Accessibility**: Restricted
- **Data Type**: CSV
- **Coverage**: Atlantic Region, Canada (including Québec), 2016–2023
- **Processing Script**: `prc_neec.R`
- **Output Files**: `neec.csv` and `substances.csv`

## Offshore petroleum activities

### Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) {#sec-offshore_petroleum_nfl}

```{r}
#| message: false
#| eval: false
offshore_petroleum_nfl <- sf::st_read("../../workspace/data/harvested/offshore_petroleum_nfl-1.0.0/processed/offshore_petroleum_nfl.gpkg", quiet = TRUE)
types <- table(offshore_petroleum_nfl$type)
```

The [Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB)](https://www.cnlopb.ca/) [@cnlopb2024] oversees and provides data on offshore petroleum activities in Newfoundland. This dataset includes multiple files covering exploration and production licenses, various well data, call for bids, land tenure sectors of the Scheduled Land Tenure System, and well info summaries (@tbl-offshore-petroleum-nfl), providing a comprehensive view of the offshore petroleum landscape in Newfoundland. The dataset includes both spatial and tabular data. Key variables encompass license IDs, well IDs, geographic coordinates, vessel types, exploration phases, and parcel identifiers. Data files are provided in various formats, including shapefiles and Excel files, and have been processed to create spatial representations of offshore petroleum activity sites. For the purpose of this project, we only retained, if available, the classification, status, start and end dates when processing the data. The different categories of offshore petroleum activities are detailed in @tbl-classifications. Each entry in the integrated dataset was given a unique identifiers. Individual entries with dual classification (e.g. exploration and delineation) were duplicated, each retaining the same id to avoid overestimating the number of entries in the dataset. As such, the number of unique entries would correspond to the number of unique identifiers rather than the number of rows in the dataset. For data with well status, we simplified status categories for the processed data, which are detailed in @tbl-status. How date ranges were established for each entry is detailed in @tbl-offshore-petroleum-nfl.


- **Source**: Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB)
- **Accessibility**: Open, Creative Commons Attribution License
- **Data Type**: Shapefiles, XLSX
- **Coverage**: Offshore regions of Newfoundland, Canada
- **Processing Script**: `prc_offshore_petroleum_nfl.R`
- **Output File**: `offshore_petroleum_nfl.gpkg`



```{r}
#| label: tbl-offshore-petroleum-nfl
#| tbl-cap: "Description of licenses, call for bids and sectors for the petroleum industry in Newfoundland and Labrador."
data.frame(
  id = c(
    "Active Exploration Licenses",
    "Active Significant Discovery Licenses",
    "Active Production Licenses",
    "Production Installations",
    "Delineation Wells",
    "Development Wells",
    "Dual Classified Wells",
    "Exploration Wells",
    "Call for Bids - Eastern Newfoundland NL23-CFB01",
    "Call for Bids - South Eastern Newfoundland NL23-CFB02",
    "Call for Bids - Eastern Newfoundland NL24-CFB01",
    "Sectors - Labrador South NL02-LS",
    "Sectors - Eastern Newfoundland NL06-EN",
    "Sectors - North Eastern Newfoundland NL01-NEN",
    "Sectors - Southern Newfoundland NL01-SN",
    "Well Info Summary"
  ),
  description = c(
    "Offshore areas in Newfoundland and Labrador that are currently licensed for petroleum exploration. An Exploration License (EL) grants rights to conduct exploratory drilling and other activities aimed at discovering petroleum resources within a designated offshore area. These licenses are typically issued for a fixed period, encouraging exploration activities that assess the resource potential in Newfoundland and Labrador's offshore regions.",
    "Offshore areas within Newfoundland and Labrador where significant petroleum discoveries have been identified and are actively licensed. An Active Significant Discovery License (SDL) allows the holder to retain rights over a discovery area with proven, potentially commercial quantities of petroleum, preserving the right to future exploration and development.",
    "Offshore areas in Newfoundland and Labrador where petroleum resources are actively being produced. A Production License (PL) authorizes the holder to extract petroleum within a specified offshore area, following proven resource discoveries. This license allows companies to conduct production activities while adhering to regulatory requirements for resource management, environmental protection, and safety.",
    "Locations of offshore production installations used in petroleum extraction activities in Newfoundland and Labrador’s offshore areas. A Production Installation refers to infrastructure such as platforms, rigs, or other facilities designed to extract, process, and transport petroleum resources from beneath the seafloor. These installations play a critical role in supporting production operations within licensed areas.",
    "Locations of delineation wells drilled in offshore areas of Newfoundland and Labrador. A Delineation Well is drilled to determine the extent and characteristics of a known petroleum discovery, helping to assess the size, quality, and commercial viability of the resource. These wells are essential in refining the understanding of hydrocarbon deposits discovered by exploration wells.",
    "Locations of development wells in offshore areas of Newfoundland and Labrador. Development Wells are drilled within proven petroleum fields to facilitate resource extraction. These wells are strategically placed to optimize production from known hydrocarbon deposits, making them critical for efficient and sustainable petroleum extraction.",
    "Offshore wells in Newfoundland and Labrador that have been classified under more than one category during their lifecycle. Dual Classified Wells may serve multiple purposes, such as transitioning from an exploration role to development or delineation, depending on the results of drilling and assessments. This dual classification reflects the adaptive approach taken in offshore petroleum exploration and production.",
    "Locations of offshore wells in Newfoundland and Labrador that are designated for exploration. Exploration Wells are drilled to investigate potential petroleum deposits in unexplored or underexplored areas. These wells are the first step in assessing hydrocarbon presence and play a crucial role in determining the viability of further development.",
    "The Call for Bids NL23-CFB01 pertains to the Eastern Newfoundland Region and was issued by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) in 2023. This call offered 28 parcels, totaling approximately 7,222,551 hectares, for exploration licensing. Interested parties were invited to submit sealed bids by November 1, 2023. However, the C-NLOPB [announced](https://www.cnlopb.ca/news/c-nlopb-announces-results-of-calls-for-bids-in-the-eastern-newfoundland-and-south-eastern-newfoundland-regions/) that no bids were received for this call.",
    "The Call for Bids NL23-CFB02 was issued by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) in 2023, offering 19 parcels totaling approximately 4,982,275 hectares in the South Eastern Newfoundland Region for exploration licensing. Interested parties were invited to submit sealed bids by November 1, 2023. However, the C-NLOPB [announced](https://www.cnlopb.ca/news/c-nlopb-announces-results-of-calls-for-bids-in-the-eastern-newfoundland-and-south-eastern-newfoundland-regions/) that no bids were received for this call.",
    "The Call for Bids NL24-CFB01 was issued by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) on April 29, 2024, inviting bids for exploration licenses in the Eastern Newfoundland Region. This call comprised 41 parcels, covering a total of 10,287,196 hectares. Interested parties were required to submit sealed bids by November 6, 2024. On November 6, 2024, the C-NLOPB [announced](https://www.cnlopb.ca/news/c-nlopb-announces-results-of-calls-for-bids-in-the-eastern-newfoundland-region/) that no bids were received for Call for Bids NL24-CFB01.",
    "The Labrador South Region (Sector NL02-LS) is an offshore area managed by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) as part of its Scheduled Land Tenure System.",
    "The Eastern Newfoundland Region (Sector NL06-EN) is an offshore area designated by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) under its Scheduled Land Tenure System.",
    "The North Eastern Newfoundland Region (Sector NL01-NEN) is an offshore area designated by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) under its Scheduled Land Tenure System.",
    "The Southern Newfoundland Region (Sector NL01-SN) is an offshore area designated by the Canada-Newfoundland and Labrador Offshore Petroleum Board (C-NLOPB) under its Scheduled Land Tenure System.",
    "Contains detailed information on wells drilled within offshore petroleum exploration and production areas. This dataset is designed to offer an inventory of wells, capturing essential data for regulatory oversight, resource management, and operational history. The dataset includes information on well names, operators, drilling dates, statuses, and geographic coordinates, among other fields."
  ),
  date_range = c(
    "The beginning is the effective date (`effective`) and the end is the second exploration period specified (`period2exp`).",
    "The beginning is the date of first exploration activity (`spud_date`) and the termination of the last renewal phase (`term_date`).",
    "The beginning is the date of first exploration activity (`spud_date`) and the termination of the last renewal phase (`term_date`).",
    "The beginning is the date of first exploration activity (`spud_date`) and the termination of the last renewal phase (`term_date`).",
    "The beginning is the date of first exploration activity (`spud_date`) and the termination of the last renewal phase (`term_date`).",
    "The beginning is the date when the well first began producing oil (`first_oil`); there is no end date specified.",
    "The beginning is the date when the date on which information about the well or license was publicly posted (`date_posted`) and the end is the specified call for bids closing year (`cfb_clse_dte`) and assumed to be on January 1st to allow for date formatting.",
    "The beginning is the date when the date on which information about the well or license was publicly posted (`date_posted`) and the end is the specified call for bids closing year (`cfb_clse_dte`) and assumed to be on January 1st to allow for date formatting.",
    "The beginning is the date when the date on which information about the well or license was publicly posted (`date_posted`) and the end is the specified call for bids closing year (`cfb_clse_dte`) and assumed to be on January 1st to allow for date formatting.",
    "The beginning is the date when the date on which information about the well or license was publicly posted (`date_posted`) and the end is the specified call for bids closing year (`cfb_clse_dte`) and assumed to be on January 1st to allow for date formatting.",
    "The beginning is the date is the opening date of the call for bids (`cfb_open_dte`) and the end is the closing date of the call for bids (`cfb_clse_dte`).",
    "The beginning is the date is the opening date of the call for bids (`cfb_open_dte`) and the end is the closing date of the call for bids (`cfb_clse_dte`).",
    "The beginning is the date is the opening date of the call for bids (`cfb_open_dte`) and the end is the closing date of the call for bids (`cfb_clse_dte`).",
    "The beginning is the date on which the sector's licensing or regulatory conditions officially took effect (`effective`); there is no end date specified.",
    "The beginning is the date on which the sector's licensing or regulatory conditions officially took effect (`effective`); there is no end date specified.",
    "The beginning corresponds to the date when drilling for the well began (`spud_date`) and the end is the date on which drilling or operational activities for the well concluded (`well_termination_date`)."
  )
) |>
  knitr::kable(col.names = c("Type", "Description", "Period"))
```



```{r}
#| label: tbl-classifications
#| tbl-cap: "Description of licenses, call for bids and sectors for the petroleum industry in Newfoundland"
data.frame(
  Classification = c(
    "Sectors",
    "Call for Bids",
    "Exploration",
    "Significant Discovery",
    "Delineation",
    "Development",
    "Production"
  ),
  Description = c(
    "Defined geographic areas designated for potential exploration and development activities, often used in the regulatory framework to organize land tenure.",
    "Competitive bidding rounds where companies submit proposals to obtain exploration or development rights for specified areas within sectors.",
    "Initial phase focused on discovering hydrocarbons, involving activities like seismic surveys and drilling exploration wells.",
    "Identification of a commercially viable hydrocarbon discovery during exploration, associated with a Significant Discovery License (SDL) to preserve rights to the area.",
    "Appraisal phase to define the size, extent, and quality of a discovered reservoir, refining understanding of the resource’s potential.",
    "Preparation phase involving construction of infrastructure, drilling of production wells, and installation of facilities needed to bring hydrocarbons to the surface.",
    "Active phase of extracting hydrocarbons, processing them, and transporting them for sale, generating revenue from the resource."
  )
) |>
  knitr::kable()
```

```{r}
#| label: tbl-status
#| tbl-cap: "Simplified Status Classifications with Descriptions"
data.frame(
  Status = c(
    "Abandoned", "Suspended", "Active Production",
    "Drilling", "Disposal", "Closed", "Off Station", NA
  ),
  Description = c(
    "Wells where activities have been permanently stopped.",
    "Wells with temporarily halted activities that may resume.",
    "Wells currently involved in production or injection activities.",
    "Wells still in the drilling phase.",
    "Wells used for waste or byproduct disposal.",
    "Wells officially closed, possibly for regulatory reasons.",
    "Wells or rigs temporarily off their regular location.",
    "No information available."
  )
) |>
  knitr::kable()
```


### Canada-Nova Scotia Offshore Petroleum Board (CNSOPB) {#sec-offshore_petroleum_ns}

```{r}
#| message: false
offshore_petroleum_ns <- sf::st_read("../../workspace/data/harvested/offshore_petroleum_ns-1.0.0/processed/offshore_petroleum_ns.gpkg", quiet = TRUE)
types <- table(offshore_petroleum_ns$type)
```

The [Canada-Nova Scotia Offshore Petroleum Board (CNSOPB)](https://www.cnsopb.ns.ca/) [@cnsopb2024] oversees and provides data relevant to offshore petroleum activities in Nova Scotia. This dataset includes multiple files covering significant discovery areas (`r types["Significant discovery areas"]` coordinates), significant discovery licenses (`r types["Significant discovery licenses"]` cooordinates), production licenses (`r types["Production licence"]` cooordinates), and the latest call for bids (`r types["Call for bids"]` polygons). The data encompasses information on various petroleum exploration and production activities. Key variables include license IDs, well IDs, geographic coordinates, exploration phases, and parcel identifiers. Data files include both Excel sheets (XLS, XLSX) and shapefiles, and they have been processed to create spatially explicit representations of petroleum activity sites. For the purpose of this project, we only retained the classification of each entry. The classification of petroleum offshore activities follows the same categories as the one presented in @tbl-classifications. Contrary to the data for Newfoudland and Labrador (see @sec-offshore_petroleum_nfl), there were no information included on the status and the time period covered.

- **Source**: Canada-Nova Scotia Offshore Petroleum Board (CNSOPB)
- **Accessibility**: Open
- **Data Type**: XLS, XLSX, Shapefiles
- **Coverage**: Offshore regions of Nova Scotia, Canada, 2009–2024
- **Processing** Script: prc_offshore_petroleum_ns.R
- **Output** File: offshore_petroleum_ns.gpkg


## Offshore wind farms

### Wind Regional Assessment - Canada {#sec-offshore_wind_can}

```{r}
#| message: false
#| eval: false
offshore_wind_can <- sf::st_read("../../workspace/data/harvested/offshore_wind_can-1.0.0/processed/offshore_wind_can.gpkg", quiet = TRUE)
```

The [Wind Regional Assessment - Canada](https://www.canada.ca/en/services/environment/conservation/assessments.html) [@wind_regional_assessment2024] identifies areas for potential offshore wind development. This dataset includes shapefiles representing:

- **Potential Future Development Areas (PFDA)** in Nova Scotia.
- **Preliminary Offshore Wind Licensing Areas (PLA)** in Newfoundland.

These areas were identified in the **Wind Regional Assessment Interim Report** for offshore wind energy. While still in the planning phase, this dataset provides critical spatial data for understanding the geographic distribution of proposed offshore wind development areas. Data from both regions are expected to be made publicly available via the Canadian Impact Assessment Registry and Open Government Portal. The data was transferred manually to the Canadian Wildlife Service and are currently stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco. For this project, data from both regions were integrated into a unified spatial layer. Each entry was given a unique identifier to track its origin and dataset of provenance (e.g., `wind_can_ns_XXXX` for Nova Scotia and `wind_can_nfl_XXXX` for Newfoundland).

- **Source**: Wind Regional Assessment - Canada
- **Accessibility**: Restricted Access
- **Data Type**: Shapefiles
- **Coverage**: Nova Scotia and Newfoundland, Canada
- **Processing Script**: `prc_offshore_wind_can.R`
- **Output File**: `offshore_wind_can.gpkg`

### Focus Area for the Regional Assessment of Offshore Wind Development in Newfoundland and Labrador {#sec-offshore_wind_nfl}

```{r}
#| message: false
#| eval: false
offshore_wind_nfl <- sf::st_read("../../workspace/data/harvested/offshore_wind_nfl-1.0.0/processed/offshore_wind_nfl.gpkg", quiet = TRUE)
```


The [Focus Area for the Regional Assessment of Offshore Wind Development in Newfoundland and Labrador](https://iaac-aeic.gc.ca/050/documents/p84343/147037E.pdf) [@iaac2024] defines priority areas for offshore wind development in Newfoundland and Labrador. The assessment, conducted by the Impact Assessment Agency of Canada, aims to provide comprehensive information, knowledge, and analysis about the potential environmental, social, economic, and health effects of offshore wind development. In November 2023, the Committee identified a **Focus Area** within a portion of the Study Area, prioritizing this region for future wind energy projects. The dataset includes spatial data representing the **Focus Area** and is accompanied by a regional assessment agreement document. Key variables include `focus_area`, `wind_energy_assessment`, and `environmental_impact`. 

- **Source**: Impact Assessment Agency of Canada
- **Accessibility**: Open Government Licence - Canada
- **Data Type**: Shapefiles, PDF
- **Coverage**: Newfoundland and Labrador, Canada
- **Processing Script**: `prc_offshore_wind_nfl.R`
- **Output File**: `offshore_wind_nfl.gpkg`


### Study Area for the Regional Assessment of Offshore Wind Development in Nova Scotia {#sec-offshore_wind_ns}

```{r}
#| message: false
#| eval: false
offshore_wind_ns <- sf::st_read("../../workspace/data/harvested/offshore_wind_ns-1.0.0/processed/offshore_wind_ns.gpkg", quiet = TRUE)
```

The [Study Area for the Regional Assessment of Offshore Wind Development in Nova Scotia](https://open.canada.ca/data/en/dataset/051c7034-c13b-49a9-9d4a-99b0b594d157) [@iaac2024b] outlines the geographic region under evaluation for future offshore wind development. This area was defined by the Impact Assessment Agency of Canada in collaboration with Natural Resources Canada and the province of Nova Scotia. The assessment focuses on evaluating the potential environmental, social, and economic effects of offshore wind energy development within the defined study area. The dataset includes spatial data for the study area and a supporting agreement document. 

- **Source**: Impact Assessment Agency of Canada
- **Accessibility**: Open Government Licence - Canada
- **Data Type**: Shapefiles, PDF
- **Coverage**: Nova Scotia, Canada
- **Processing Script**: `prc_offshore_wind_ns.R`
- **Output File**: `offshore_wind_ns.gpkg`

### Bureau of Ocean Energy Management (BOEM) Offshore Wind Data {#sec-offshore_wind_usa}

```{r}
#| message: false
#| eval: false
offshore_wind_usa <- sf::st_read("../../workspace/data/harvested/offshore_wind_ns-1.0.0/processed/offshore_wind_usa.gpkg", quiet = TRUE)
```

The [Bureau of Ocean Energy Management (BOEM)](https://www.boem.gov/) [@boem2024] manages offshore wind energy activities across U.S. waters. This dataset provides detailed geospatial and tabular data on wind energy development, including lease areas, planning areas, marine hydrokinetic leases, project phase areas, and infrastructure components such as turbines, substations, and cable systems (@tbl-wind-usa-classification). Each entry in the dataset is accompanied with a lease or protraction number (see @tbl-wind-usa-number). The dataset supports decision-making and analysis for renewable energy development by offering comprehensive coverage of offshore wind projects. Access to the data requires a ArcGIS online membership. Data were therefore harvested and stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco.

- **Source**: Bureau of Ocean Energy Management (BOEM)
- **Accessibility**: Open Government Licence - United States
- **Data Type**: Shapefiles, Geodatabase
- **Coverage**: United States Offshore, 2020–2024
- **Processing Script**: `prc_offshore_wind_usa.R`
- **Output File**: `offshore_wind_usa.gpkg`


```{r}
#| label: tbl-wind-usa-classification
#| tbl-cap: "Classifications and descriptions of BOEM offshore wind data components."
data.frame(
  Classification = c(
    "Leases",
    "Planning areas",
    "MHK leases and planning areas",
    "Project phase areas",
    "Substations",
    "Turbine locations",
    "Cable interconnections",
    "Ocean observing devices",
    "Export cable corridors",
    "Cable landings",
    "Project inter-array cables"
  ),
  Description = c(
    "Represents areas designated for offshore wind energy activities under BOEM's leasing program. Includes four types of leases: 'Commercial' leases for large-scale wind energy production, 'Easement' leases for infrastructure such as cables or substations, 'Right-of-Way Grant' leases for energy transmission corridors, and 'Research' leases for testing and studying renewable energy technologies.",
    "Defines geographic areas identified by BOEM for potential offshore wind development. These areas are used for assessing wind energy resources and determining where to initiate leasing activities, based on environmental, economic, and technical considerations.",
    "Includes Marine Hydrokinetic (MHK) areas for renewable energy from ocean-based resources such as waves, tides, or currents. MHK leases encompass 'Research' leases for testing technology, 'Right-of-Way Grant' leases for transmission infrastructure, and 'Easement Area' leases for supporting installations like substations.",
    "Represents specific geographic areas proposed for different phases of offshore wind energy projects. These areas are identified based on project development stages, including site assessment, construction, and operations.",
    "Provides point data representing substations with their status (either 'installed' or 'proposed') and location type (either 'offshore' or 'onshore').",
    "Provides point data representing the locations of wind turbines with their status as either 'proposed' or 'installed'.",
    "Represents spatial data for cable interconnections linking offshore wind installations to onshore infrastructure or between offshore installations. Includes the status of cables as either 'proposed' or 'installed'.",
    "Includes point data for devices deployed to observe ocean conditions in areas relevant to offshore wind development. Observing devices monitor factors such as currents, waves, and water quality, and are categorized by their status as either 'proposed' or 'installed'.",
    "Represents spatial data for corridors used by export cables connecting offshore wind farms to onshore substations. Includes the status of corridors as either 'proposed' or 'installed'.",
    "Provides point data for locations where export cables come ashore to connect offshore wind farms to the terrestrial power grid. Includes the status of cable landings as either 'proposed' or 'installed'.",
    "Represents spatial data for inter-array cables within an offshore wind farm. These cables connect turbines to each other and to the offshore substation, with their status categorized as either 'proposed' or 'installed'."
  )
) |>
  knitr::kable()
```


```{r}
#| label: tbl-wind-usa-number
#| tbl-cap: "Comparison of lease numbers and protraction numbers in BOEM offshore wind datasets"
data.frame(
  Aspect = c(
    "Purpose",
    "Format",
    "Focus",
    "Context"
  ),
  Lease_Numbers = c(
    "Identifies specific lease agreements for offshore activities.",
    "Often alphanumeric (e.g., 'OCS-A 0501').",
    "Administrative and contractual.",
    "Specific to a lessee's activities."
  ),
  Protraction_Numbers = c(
    "Identifies geographic areas on the Outer Continental Shelf (OCS) used for spatial organization.",
    "Often numeric or grid-based (e.g., 'NG14-02').",
    "Geographic and spatial organization.",
    "Applicable to the broader mapping system."
  )
) |>
  knitr::kable(col.names = c("", "Lease numbers", "Protraction numbers"))
```


## Automatic Identification System (AIS) Shipping Data {#sec-shipping_ais}

The [Automatic Identification System (AIS)](https://tc.canada.ca/en/marine-transportation/navigation-aids/automatic-identification-system-ais) [@tc2023] collects dynamic positional data from ships to enhance maritime safety, monitor environmental impacts, and facilitate shipping analyses. This dataset, provided by Transport Canada, focuses on 2023 vessel movements across Atlantic Canada. The AIS data provides dynamic vessel information, including position, speed, and course, captured through terrestrial, satellite, and tier-based receivers (@tbl-shipping-ais-attr). The AIS data shared includes both terrestrial and satellite AIS messages, fused to minimize spatial and temporal gaps and stored as compressed parquet files.The data shared also contains pre-processed tracklines as a geodatabase. However, we used the raw AIS positional data to recreate tracklines for the purposes of this project to allow for more temporal granularity. The data was transferred manually to the Canadian Wildlife Service and are currently stored on a secure Google Cloud Storage bucket accessible through an authentication key and maintained by inSileco.

- **Source**: Transport Canada
- **Accessibility**: Restricted
- **Data Type**: Parquet, Geodatabase
- **Coverage**: Atlantic Canada, 2023
- **Processing Script**: `prc_shipping_ais_tracklines.R`
- **Output File**: `shipping_ais_tracklines.gpkg`


```{r}
#| label: tbl-shipping-ais-attr
#| tbl-cap: "Columns (attributes) associated with the derived positional AIS parquet files (table from @tc2023)."

data.frame(
  Column_Name = c("MMSI", "Latitude", "Longitude", "SOG", "COG", "Heading", "YMD_HMS", "Source"),
  Description = c(
    "Maritime Mobile Service Identity",
    "Latitude of the positional message",
    "Longitude of the positional message",
    "Speed over ground of vessel",
    "Course over ground of vessel",
    "True heading of vessel",
    "Time of positional message in UTC",
    "Data source/message type of AIS message"
  ),
  Information = c(
    "9-digits (MMSIs with valid ship country codes are from 201000000 to 775999999). All ship MMSIs use the format M1I2D3X4X5X6X7X8X9, the first three digits represent the Maritime Identification Digits (MID) and they represent the flag/nationality of the ship. The following digits (X4-X9) can be any numerical value from 0 to 9.",
    "Latitude recorded in decimal degrees with an accuracy of approximately 10 metres (-90 to 90).",
    "Longitude recorded in decimal degrees with an accuracy of approximately 10 metres (-180 to 180).",
    "The speed of the ship with respect to the ground, recorded in knots (0 to 102.2 are valid; 102.3 is invalid).",
    "The course of the ship, recorded in degrees (0-359.9 are valid; 360 is invalid).",
    "The true heading of the ship, recorded in degrees (0-359 are valid; 360 is invalid).",
    "The UTC time stamp of when the positional AIS message was sent from the ship (YYYY-MM-DD hh:mm:ss).",
    "The source (e.g., terrestrial, dynamic, satellite) and class (A or B) of the AIS message sent (e.g., TAIS_A, TAIS_B, SAIS_A)."
  ),
  Data_Type = c(
    "Integer", "Float", "Float", "Float", "Float", "Integer", "Datetime", "String"
  )
) |>
  knitr::kable()
```


### Data Processing Overview
To create meaningful tracklines representing vessel movements over space and time, we followed the processing steps described in @veinot2023:

1. **Data Extraction**
   - AIS data was extracted from compressed archive files and loaded into memory using efficient tools such as the `arrow` and `vroom` R packages. This ensured scalability given the dataset size, which often exceeds 10GB annually.
2. **Initial Data Cleaning**
   - Raw AIS data was filtered to remove invalid records:
     - Only positions with valid **MMSI** values (201000000–775999999) were retained.
     - **SOG** values were restricted to the range 0.1–100 knots to remove implausible speeds.
3. **Day/Night Classification**
   - Using the `suncalc` R package, each AIS position was classified as daytime or nighttime. The classification was based on local sunrise and sunset times at each position, ensuring alignment with the UTC timestamp.
4. **Trackline Generation**
   - Tracklines were created by:
     1. Sorting AIS data by **MMSI** and timestamp.
     2. Calculating distances between consecutive positions using geodesic measurements.
     3. Splitting movements into separate tracks when:
        - The distance exceeded 50 nautical miles.
        - The time interval between positions exceeded 300 minutes.
     4. Further segmenting tracklines to differentiate between daytime and nighttime navigation.
5. **Trackline Attributes**
   - Additional attributes were calculated for each trackline to facilitate further analysis:
     - **Average and Maximum Speeds**: Derived from AIS speed data.
     - **Number of Positions**: Total AIS messages contributing to the trackline.
     - **Elapsed Hours**: Duration of the trackline.
     - **Track Length**: Computed in kilometers using geodesic measurements.
     - **Daytime Classification**: Indicates whether the trackline occurred during the day, night, or a mix of both.
6. **Post-Processing**
   - Erratic tracklines were flagged and removed using heuristic thresholds:
     - Extremely short tracks (<100 meters).
     - Unrealistic average speeds (<0.3 knots or >100 km/h).
     - Speeds exceeding vessel-type-specific thresholds:
      - 30 km/h for pleasure vessels and special ships.
      - 35 km/h for fishing and tug vessels.
      - 40 km/h for all other vessel types.
   - Tracklines identified as spoofing were excluded from further analyses.
7. **Output**
   - The cleaned tracklines were exported as a Geopackage file (`.gpkg`) Each trackline includes spatial geometry and relevant attributes, such as speed, duration, and daytime classification, enabling detailed analyses of vessel movements across Canadian waters (@tbl-shipping-ais-tracklines).

```{r}
#| label: tbl-shipping-ais-tracklines
#| tbl-cap: "Columns (attributes) associated with the tracklines created from the derived positional AIS parquet files."
data.frame(
  Column_Name = c(
    "mmsi", "track_id", "day_night_segment", "start_time", "end_time", "avg_sog",
    "max_sog", "n_pos", "elapsed_hours", "month", "year", "daytime", "geometry",
    "track_length_km", "vess_name", "ntype", "specific_vess_type", "mt_gt",
    "mt_length", "mt_width", "mt_draught_max", "GT_ID_TC", "avg_speed_kmh", "spoofing_flags"
  ),
  Description = c(
    "Maritime Mobile Service Identity",
    "Unique identifier for each trackline within a ship's movement",
    "Unique identifier for segments of tracklines split by day or night",
    "Start time of the trackline",
    "End time of the trackline",
    "Average speed over ground for the trackline",
    "Maximum speed over ground for the trackline",
    "Number of AIS positional records in the trackline",
    "Total elapsed hours for the trackline",
    "Month in which the trackline occurs",
    "Year in which the trackline occurs",
    "Classification of trackline navigation as day, night, or mixed",
    "Geospatial representation of the trackline as a linestring",
    "Length of the trackline in kilometers",
    "Vessel name",
    "General vessel type",
    "Specific vessel type",
    "Gross tonnage of the vessel",
    "Length of the vessel in meters",
    "Width of the vessel in meters",
    "Maximum draught of the vessel in meters",
    "Gross tonnage classification based on predefined intervals",
    "Average speed of the vessel during the trackline in kilometers per hour",
    "Flag indicating whether the trackline is suspected of spoofing"
  ),
  Information = c(
    "9-digit identifier assigned to a vessel",
    "Calculated based on movement and distance thresholds",
    "Generated to separate trackline segments into day and night",
    "Recorded in POSIXct format with timezone UTC",
    "Recorded in POSIXct format with timezone UTC",
    "Measured in knots converted to kilometers per hour",
    "Measured in knots converted to kilometers per hour",
    "Count of AIS points within the trackline",
    "Calculated from start and end timestamps",
    "Numerical month value (1-12)",
    "Numerical year value (e.g., 2023)",
    "Derived using sunrise and sunset times",
    "LINESTRING geometry object in CRS EPSG:4326",
    "Derived from geospatial data and measured in kilometers",
    "Obtained from the AIS static data",
    "Categorized based on the AIS static data",
    "Detailed classification provided in AIS static data",
    "From AIS static data; corresponds to ship's gross tonnage",
    "From AIS static data; corresponds to ship's length",
    "From AIS static data; corresponds to ship's width",
    "From AIS static data; corresponds to ship's draught",
    "Categorization of gross tonnage into predefined bins",
    "Calculated by dividing track length by elapsed hours",
    "Binary value indicating if trackline is potentially spoofed"
  ),
  Data_Type = c(
    "Numeric", "Integer", "Integer", "POSIXct", "POSIXct", "Numeric",
    "Numeric", "Integer", "Units [hours]", "Numeric", "Numeric", "Character",
    "sfc_LINESTRING", "Units [km]", "Character", "Character", "Character",
    "Numeric", "Numeric", "Numeric", "Numeric", "Character", "Units [km/h]",
    "Logical"
  )
) |>
  knitr::kable()
```

## Visible Infrared Imaging Radiometer Suite 

The Visible Infrared Imaging Radiometer Suite (VIIRS) is a key instrument aboard the National Oceanic and Atmospheric Administration's (NOAA) Suomi National Polar-orbiting Partnership (Suomi NPP) and NOAA-20 satellites. VIIRS collects visible and infrared imagery and radiometric measurements, supporting applications such as environmental monitoring, weather forecasting, and maritime surveillance, including boat detection.

### VIIRS Boat Detection (VBD) Dataset {#sec-viirs_boat_detection}

The [VIIRS Boat Detection (VBD) Dataset](https://eogdata.mines.edu/products/vbd/) [@eog2024] provides nightly global data on boat detections based on radiance observations captured by the Visible Infrared Imaging Radiometer Suite (VIIRS). This dataset is widely used for maritime surveillance, tracking vessel activity, and monitoring illegal fishing. The raw data includes detections with associated radiance, confidence metrics, and spatial coordinates (@tbl-viirs-vbd-attr). The data was acquired for the years 2020–2023 and pre-processed to facilitate further analysis of vessel patterns within the Canadian Exclusive Economic Zone (EEZ). The processed data is stored as a Geopackage and includes filtered, high-confidence detections with key attributes for radiance intensity and spatial location.

- **Source**: Earth Observation Group, Payne Institute for Public Policy
- **Accessibility**: Public domain with registration
- **Data Type**: CSV files
- **Coverage**: Global, 2020–2023
- **Processing Script**: `prc_viirs_boat_detection.R`
- **Output File**: `viirs_boat_detection.gpkg`

```{r}
#| label: tbl-viirs-vbd-attr
#| tbl-cap: "Columns (attributes) associated with the VIIRS Boat Detection dataset used for this project."

data.frame(
  Column_Name = c("Date_Mscan", "Lat_DNB", "Lon_DNB", "Rad_DNB", "QF_Detect"),
  Description = c(
    "Date and time of the observation in UTC",
    "Latitude of the detected boat in decimal degrees",
    "Longitude of the detected boat in decimal degrees",
    "Radiance intensity detected at the location (nW/cm²/sr)",
    "Quality flag indicating confidence in the detection (1 = high confidence)"
  ),
  Data_Type = c("POSIXct", "Numeric", "Numeric", "Numeric", "Integer")
) |>
  knitr::kable()
```

#### Data Processing Overview

To prepare the VBD dataset for further analyses, the following pre-processing steps were applied:

- Raw `.csv.gz` files were downloaded and processed using the `arrow` package, which allows for efficient reading of compressed CSV files.
- High-confidence detections were retained by filtering for rows with:
  - **QF_Detect = 1**: Ensures only high-confidence detections are included.
  - **Rad_DNB > 0**: Removes detections with non-positive radiance values.
- Optionally, filters for **SMI** (Signal-to-Mean Index) and **SHI** (Signal-to-High Index) can be applied for stricter data quality criteria (e.g., **SMI > 0.1**, **SHI > 0.75**).
- The filtered dataset was converted to a spatial object with spatial projection **EPSG:4326** using the `sf` package, enabling geospatial analyses and integration with other datasets.
- Observation timestamps were standardized to UTC using the `lubridate` package to ensure consistent temporal referencing across datasets.
- The pre-processed dataset was saved as a Geopackage (`viirs_boat_detection.gpkg`)


#### Signal-to-Mean Index (SMI) and Signal-to-High Index (SHI) in VIIRS Boat Detection Data

The **Signal-to-Mean Index (SMI)** and **Signal-to-High Index (SHI)** are metrics included in the VBD dataset that provide additional quality indicators for radiance detections. While these indices were not directly applied in the current processing pipeline, they could be used for further refining data quality and reducing false positives:

- **SMI**: Measures the signal strength relative to the mean radiance value in the surrounding area. Higher values indicate stronger, more distinct detections.
- **SHI**: Measures the signal strength relative to a high radiance threshold, helping to exclude detections influenced by bright land-based lights or other high-radiance sources.

Future iterations of the pipeline could incorporate these metrics for scenarios requiring stricter detection thresholds, such as high-confidence studies of vessel activity in heavily lit areas.




## References